---
title: "Landcover Classification for Santa Barbara County"
description: "Utilizing Landsat data for Santa Barbara County to train and apply a decision tree classifier for landcover types"
author: "Carly  Caswell"
date: "2023-23-11"
output:
    html_document:
      print_df: paged
      toc: yes
      toc_depth: 4
      toc_float: yes
---


# Overview

Monitoring the distribution and change in land cover types can help us understand the impacts of phenomena like climate change, natural disasters, deforestation, and urbanization. Determining land cover types over large areas is a major application of remote sensing because we are able to distinguish different materials based on their spectral reflectance.

Classifying remotely sensed imagery into landcover classes enables us to understand the distribution and change in landcover types over large areas. I am going to perform a landcover classification using a supervised approach; a decision tree classifier. More specifically, I'm going to complete a land cover classification for southern Santa Barbara County based on multi-spectral imagery and data on the location of 4 land cover types: green vegetation, dry grass or soil, urban, and water

## Goals

1. Load and process Landsat scene
2. Crop and mask Landsat data to study area
3. Extract spectral data at training sites
4. Train and apply decision tree classifier
5. Plot the results

## Data

**Landsat 5 Thematic Mapper**\

-   [Landsat 5](https://www.usgs.gov/landsat-missions/landsat-5)
-   1 scene from September 25, 2007\
-   bands: 1, 2, 3, 4, 5, 7
-   Collection 2 surface reflectance product\

**Study area and training data**

-   polygon representing southern Santa Barbara county
-   polygons representing training sites\
    -   type: character string with land cover type\

# My Workflow

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing data

#### Loading packages and setting my working directory

```{r include=TRUE, message=FALSE, warning=FALSE}
library(sf)
library(terra)
library(here)
library(dplyr)
library(rpart)
library(rpart.plot) #for training and classifying aglorithm
library(tmap)

rm(list = ls())

#Setting working directory:
here::i_am("landcover-classification/landcover-classification.Rmd")
setwd(here())
```

#### Importing Landsat Data

First, I'm going to create a raster stack based on the 6 bands I plan to work with. I will create a raster stack with the list of files and will then update the names of the layers to match the spectral bands.

```{r include=TRUE}
#Listing files for each band, including the full file path
filelist <- list.files("./data/landsat-data/", full.names = TRUE)

#Reading in and storing files as a raster stack
landsat_20070925 <- rast(filelist)

#Updating layer names to match band
names(landsat_20070925) <- c("blue", "green", "red", "NIR", "SWIR1", "SWIR2")

#Plotting a true color image to check out the Landsat data
plotRGB(landsat_20070925, r = 3, g = 2, b = 1, stretch = "lin")
```

#### Loading My Study Area

I want to limit my analysis to the southern portion of Santa Barbara County where I have training data, so I'll read in a file that defines the area I plan to use.

```{r include=TRUE}
# Reading in shapefile for southern portion of SB county
SB_county_south <- st_read("./data/week9/SB_county_south.shp")

#Projecting to match Landsat data crs
SB_county_south <- st_transform(SB_county_south, crs = crs(landsat_20070925))
```

#### Cropping and masking Landsat data to my study area

Now, I can crop and mask the Landsat data to the study area. This will reduce the amount of data I'lll be working with and will save on computational time. 

```{r include=TRUE}
# Cropping Landsat scene to the extent of the SB county shapefile
landsat_cropped <- crop(landsat_20070925, SB_county_south)

# Masking the raster to southern portion of SB county
landsat_masked <- mask(landsat_cropped, SB_county_south)

# Removing unnecessary objects from my environment
rm(landsat_20070925, SB_county_south, landsat_cropped)
```

#### Converting Landsat Values to Reflectance

I need to convert the values in my raster stack to correspond to reflectance values. To do this I'm going to need to remove erroneous values and apply any scaling factors to convert to reflectance.

Because I am working with [Landsat Collection 2](https://www.usgs.gov/landsat-missions/landsat-collection-2), the valid range of pixel values for this collection is 7,273-43,636, with a multiplicative scale factor of 0.0000275 and an additive scale factor of -0.2. So, I'm going to reclassify any erroneous values as `NA` and update the values for each pixel based on the scaling factors. Once done, I will have pixel values ranging from 0-100% which will allow me to complete an image classification.

```{r include=TRUE}
# reclassify erroneous values as NA
rcl <- matrix(c(-Inf, 7273, NA,
                 43636, Inf, NA), ncol = 3, byrow = TRUE)

landsat <- classify(landsat_masked, rcl = rcl)

# adjusting values based on the scaling factor
landsat <- (landsat * 0.0000275 - 0.2) * 100

# plotting the true color image to check my results
plotRGB(landsat, r = 3, g = 2, b = 1, stretch = "lin")

# checking that my values are 0 - 100
summary(landsat)
```

## Classifying My Image

#### Extracting Reflectance Values for Training Data

I'm going to load the shapefile identifying different locations within the study area as containing one of the 4 land cover types. I am then going to extract the spectral values at each site to create a data frame that relates land cover types to their spectral reflectance.

```{r include=TRUE}
# reading in and transforming training data
training_data <- st_read("./data/week9/trainingdata.shp") %>%
  st_transform(., crs = crs(landsat))

# extracting reflectance values at training sites
training_data_values <- extract(landsat, training_data, df = TRUE)

# converting training data to data frame
training_data_attributes <- training_data %>%
  st_drop_geometry()

# joining training data attributes and extracted reflectance values
SB_training_data <- left_join(training_data_values, training_data_attributes,
                              by = c("ID" = "id")) %>%
  mutate(type = as.factor(type)) # convert landcover type to factor
```

#### Training Decision Tree Classifier

To train my decision tree, I'm going to establish a model formula (i.e. what the response and predictor variables are). I will use the `rpart` function which is going to implement the [CART algorithm](https://medium.com/geekculture/decision-trees-with-cart-algorithm-7e179acee8ff). 

To understand how the decision tree classifies pixels, I'm then going to plot my results. 

```{r include=TRUE}
#Establishing an initial formula
SB_formula <- type ~ red + green + blue + NIR + SWIR1 + SWIR2

#Training the decision tree on my training data
SB_decision_tree <- rpart(formula = SB_formula,
                          data = SB_training_data,
                          method = "class",
                          na.action = na.omit)

#Plotting the decision tree to understand the classification
prp(SB_decision_tree)
```

#### Applying a Decision Tree

Now that I created a decision tree, I need to apply it to the entire Landsat image. I will use the `terra` package, more specifically the `predict()` function, which will allow me to apply a model to the data. The `predict()` function will return a raster layer with integer values and those will correspond to the *factor levels* in the training data. I will check the levels of the training data to understand each integers correspondence to the categories. 

```{r include=TRUE}
#Classifying the image based on my decision tree
SB_classification <- predict(landsat, SB_decision_tree, type = "class", na.rm = TRUE)

#Checking the level to understand the order of classes in the prediction
levels(SB_training_data$type)

```

#### Plotting Results

Now I can plot to see my land cover map of Santa Barbara!

```{r}
tm_shape(SB_classification) +
  tm_raster(col.scale = tm_scale_categorical(values = c("#8DB580", "#F2DDA4", "#7E8987", "#6A8EAE")),
            col.legend = tm_legend(labels = c("green vegetation", "soil/dead grass", "urban", "water"),
                                   title = "Landcover type")) +
  tm_layout(legend.position = c("left", "bottom"))
```
